Kalman tracking and estimation of continuous gravitational waves with a pulsar timing array: inclusion of the pulsar terms

### 2023-10-11 manuscript/manuscript_2/paper.tex
PP plots, beta

\subsection{Challenges of pulsar term}
\begin{figure*}
	\includegraphics[width=\textwidth, height =0.5\textwidth ]{images/Likelihoods_alpha}
	\caption{abc}
	\label{fig:likelihood_comparison}
\end{figure*}

K23 uses a likelihood-based Bayesian framework for inference of the model parameters and model selection. That is, given a temporal sequence of noisy measurements $\boldsymbol{Y}$, the posterior distribution of $\boldsymbol{\theta}$ is calculated by 
Bayes' Rule,
\begin{equation}
	p(\boldsymbol{\theta} | \boldsymbol{Y}) = \frac{\mathcal{L}(\boldsymbol{Y} | \boldsymbol{\theta}) \cdot \pi(\boldsymbol{\theta})}{\mathcal{Z}} \ ,
\end{equation}
where $\pi(\boldsymbol{\theta})$ is the prior distribution on $\boldsymbol{\theta}$ and $\mathcal{Z}$ is the marginalised likelihood, or evidence
\begin{equation}
	\mathcal{Z} = \int d \boldsymbol{\theta} \mathcal{L}(\boldsymbol{Y} | \boldsymbol{\theta})  \pi(\boldsymbol{\theta})  \ . \label{eq:model_evidence}
\end{equation}
and $\mathcal{L}(\boldsymbol{Y}| \boldsymbol{\theta})$ is the likelihood calculated via a Kalman filter \citep{Kalman1}. Nested sampling \citep{Skilling} is used to estimate $p(\boldsymbol{\theta} | \boldsymbol{Y})$ and $\mathcal{Z}$. Given a specific realisation of $\boldsymbol{Y}$, $\mathcal{L}(\boldsymbol{Y}| \boldsymbol{\theta})$ is solely a function of $\boldsymbol{\theta}$, i.e. $\mathcal{L}(\boldsymbol{Y}| \boldsymbol{\theta}) = \mathcal{L}(\boldsymbol{\theta})$. We will use the notation $\mathcal{L}(\Theta)$ to refer to the likelihood returned by the Kalman filter for parameter $\Theta$, where $\Theta \in \boldsymbol{\theta}$ and all other members of $\boldsymbol{\theta}$ are held constant at their true injected values. \newline

To elucidate the complications introduced by including the pulsar term, we generate some noisy synthetic data $\boldsymbol{Y}$ using the methods outlined in K23. $\boldsymbol{\theta}_{\rm gw}$ are chosen arbitrarily, with $\alpha = 1.0 $ rad.
The choice of $\boldsymbol{\theta}_{\rm psr}$ corresponds to the 47 pulsars in the 12.5 year NANOGrav data set  \citep{2020ApJ...905L..34A}. We run the Kalman filter using the measurement equation inclusive of the pulsar terms, Equation \eqref{eq:g_func_trig} on $\boldsymbol{Y}$ for different values of $\alpha$, $ 0.0 \le \alpha \le \pi$ rad. We obtain a curve $\mathcal{L}(\alpha)$, which is plotted in the left hand panel of Figure \ref{fig:likelihood_comparison} (blue curve). The dashed grey line indicates the injected value of $\alpha$. The inset plot shows a zoomed-in section of $\mathcal{L}(\alpha)$ within the region $ 1.0 - 10^{-3} \le \alpha \le 1.0 +10^{-3}$ rad. There are two important features in the figure. The first is that on the scale of radians the curve of $\mathcal{L}(\alpha)$ is very noisy. The second is that despite being noisy, a true likelihood maximum coincident with the injected value does exist.

This means that whilst it is theoretically possible to use likelihood-based methods for

practically the noise in the curve

Why noise is bad

Similar are observed for $\Omega$ and $\delta$, the other members of $\theta_{\rm gw}$ are smooth and concave.

$\mathcal{L}(\Theta)$ is noisy

a Gauss-Markov model used to algorithmically recover a temporal sequence of stochastically evolving  system state variables, $\boldsymbol{X}(t)$, which are not observed directly, given a temporal sequence of noisy measurements, $\boldsymbol{Y}(t)$. It finds common use in engineering applications and has been applied successfully in neutron star astrophysics \citep[e.g.][]{Myers2021MNRAS.502.3113M,Meyers2021,Melatos2023}.

$p(\boldsymbol{\theta} | \boldsymbol{Y})$ is calculated via nested sampling

@@

\subsection{New parameterisation}

To alleviate the probelsm of seciton X,

>>>>

\subsection{Challenges of pulsar term}\label{sec:pulsar_terms_challenge}
\begin{figure*}
	\includegraphics[width=\textwidth, height =0.5\textwidth ]{images/Likelihoods_alpha}
	\caption{abc}
	\label{fig:likelihood_comparison}
\end{figure*}

K23 uses a likelihood-based Bayesian framework for inference of the model parameters and model selection. That is, given a temporal sequence of noisy measurements $\boldsymbol{Y}$, the posterior distribution of $\boldsymbol{\theta}$ is calculated by 
Bayes' Rule,
\begin{equation}
	p(\boldsymbol{\theta} | \boldsymbol{Y}) = \frac{\mathcal{L}(\boldsymbol{Y} | \boldsymbol{\theta}) \cdot \pi(\boldsymbol{\theta})}{\mathcal{Z}} \ ,
\end{equation}
where $\pi(\boldsymbol{\theta})$ is the prior distribution on $\boldsymbol{\theta}$ and $\mathcal{Z}$ is the marginalised likelihood, or evidence
\begin{equation}
	\mathcal{Z} = \int d \boldsymbol{\theta} \mathcal{L}(\boldsymbol{Y} | \boldsymbol{\theta})  \pi(\boldsymbol{\theta})  \ . \label{eq:model_evidence}
\end{equation}
and $\mathcal{L}(\boldsymbol{Y}| \boldsymbol{\theta})$ is the likelihood calculated via a Kalman filter \citep{Kalman1}. Nested sampling \citep{Skilling} is used to estimate $p(\boldsymbol{\theta} | \boldsymbol{Y})$ and $\mathcal{Z}$. Given a specific realisation of $\boldsymbol{Y}$, $\mathcal{L}(\boldsymbol{Y}| \boldsymbol{\theta})$ is solely a function of $\boldsymbol{\theta}$, i.e. $\mathcal{L}(\boldsymbol{Y}| \boldsymbol{\theta}) = \mathcal{L}(\boldsymbol{\theta})$. We will use the notation ONE DIMENSIONAL $\mathcal{L}(\Theta)$ to refer to the likelihood returned by the Kalman filter for parameter $\Theta$, where $\Theta \in \boldsymbol{\theta}$ and all other members of $\boldsymbol{\theta}$ are held constant at their true injected values. \newline

To elucidate the complications introduced by including the pulsar term, we generate some noisy synthetic data $\boldsymbol{Y}$ using the methods outlined in K23. $\boldsymbol{\theta}_{\rm gw}$ are chosen arbitrarily, with $\alpha = 1.0 $ rad.
The choice of $\boldsymbol{\theta}_{\rm psr}$ corresponds to the 47 pulsars in the 12.5 year NANOGrav data set  \citep{2020ApJ...905L..34A}. We run the Kalman filter using the measurement equation inclusive of the pulsar terms, Equation \eqref{eq:g_func_trig} on $\boldsymbol{Y}$ for different values of $\alpha$, $ 0.0 \le \alpha \le \pi$ rad. We obtain a curve $\mathcal{L}(\alpha)$, which is plotted in the left hand panel of Figure \ref{fig:likelihood_comparison} (blue curve). The dashed grey line indicates the injected value of $\alpha$. The inset plot shows a zoomed-in section of $\mathcal{L}(\alpha)$ within the region $ 1.0 - 10^{-3} \le \alpha \le 1.0 +10^{-3}$ rad. There are two important features in the figure.

The first is that on scales of order the width of the prior $\pi(\alpha)$ (i.e radians), the curve of $\mathcal{L}(\alpha)$ is very noisy. The second is that despite being noisy, a true likelihood maximum coincident with the injected value does exist and is locally concave on scales $\mathcal{O} \left( 10^{-4}\right)$ radians.

This means that whilst it is theoretically possible to use likelihood-based methods for Bayesian estimation of $\alpha$, in practice the computation is intractable since the noise in the likelihood curve inhibits sampling algorithms to reliably converge to a definitive posterior.

Naturally $\mathcal{L}(\alpha)$ is only one dimensional. The situation is further esacerbated when we move to $N +D $ dimensions.

However, the majority of parameters are well-behaved and exhibit smooth concanve 1D

It is just omega, delta and alpha

Why these parameters?

These are the parameters which correspond to the pulsar term

go through N cycles

Not all parameters are noisy

Whilst noisy likelihood inference is possible...

@@

\subsection{New parametrisation}
In this section we introduce a new parametrisation of the model described in Section \ref{sec:model} in order to alleviate the problems outlined in Section \ref{sec:pulsar_terms_challenge}. \newline
